---
title: "Project report"
author: "Mario Limón"
date: "2025-02-02"
output: 
  pdf_document:
    toc: yes
    toc_depth: 3
bibliography: "./Proyect_references.bib"
---
## Data process

#### Libraries
```{r}
library("recount3")
library("ggplot2")
```

#### Download data
Downloading data from recount 3 database of project 'Transcriptome profiling in
knock-in mouse models of Huntington''s disease  [Striatum_mRNA] (SRP053398). The
data of this study is to analyze the expression profiles of RNA in mouse models 
with Hungtington's diseaase at different ages with different lengths of CAG 
repeats, to understand better the molecular changes related with this neurodegenarative 
disease.


```{r}
rse_gene_SRP053398 <- recount3::create_rse_manual(
  project = "SRP053398",
  project_home = "data_sources/sra",
  organism = "mouse",
  annotation = "gencode_v23",
  type = "gene")
```
#### Processing data

```{r}
# Transform the counts
assay(rse_gene_SRP053398, "counts") <- compute_read_counts(rse_gene_SRP053398)

# Before expanding the data we should explore the attributes
rse_gene_SRP053398$sra.sample_attributes[1:3]
```

```{r}
# Now we expand the attributes to process the data to make easier the analysis
rse_gene_SRP053398 <- expand_sra_attributes(rse_gene_SRP053398)
colData(rse_gene_SRP053398)[
    ,
    grepl("^sra_attribute", colnames(colData(rse_gene_SRP053398)))]
```


```{r}
# To perform the statistical analysis we need to use a correct R format
rse_gene_SRP053398$sra_attribute.age <- as.numeric(gsub(" month", "", rse_gene_SRP053398$sra_attribute.age))
rse_gene_SRP053398$sra_attribute.genotype <- factor(tolower(rse_gene_SRP053398$sra_attribute.genotype))
rse_gene_SRP053398$sra_attribute.Sex <- factor(rse_gene_SRP053398$sra_attribute.Sex)

# We can check the summary of the interest variables
summary(as.data.frame(colData(rse_gene_SRP053398)[
    ,
    grepl("^sra_attribute.[age|disease|RIN|sex]", colnames(colData(rse_gene_SRP053398)))
]))
```



