---
title: "Project report"
author: "Mario Limón"
date: "2025-02-02"
output: 
  pdf_document:
    toc: yes
    toc_depth: 3
bibliography: "./Proyect_references.bib"
---
## Data process

#### Libraries
```{r}
library("recount3")
library("ggplot2")
```

#### Download data
Downloading data from recount 3 database of project 'Chracterization_of_iPSC_der
ived_macrophages___cardiovascular_pilot_Open_access'. This study tries to understand
the genetic mechanisms behind the formation of foam cells in the atherosclerosis
pathology, using iPSC-derived macrophages, exposed to acetylated LDL (acLDL). 

```{r}
rse_data_ERP022909 <- recount3::create_rse_manual(
  project = "ERP022909",
  project_home = "data_sources/sra",
  organism = "human",
  annotation = "gencode_v26",
  type = "gene")
```
#### Processing data

```{r}
# Transform the counts
assay(rse_data_ERP022909, "counts") <- compute_read_counts(rse_data_ERP022909)

# Before expanding the data we should explore the attributes
rse_data_ERP022909$sra.sample_attributes[1:3]
```




