---
title: "Project report"
author: "Mario Limón"
date: "2025-02-02"
output: 
  pdf_document:
    toc: yes
    toc_depth: 3
bibliography: "../Proyect_references.bib"
---
## Data process

#### Libraries
```{r}
library("recount3")
library("ggplot2")
library("edgeR")
library("limma")
library("pheatmap")
library("RColorBrewer")
```

#### Download data
Downloading data from recount 3 database of project 'Transcriptome profiling in
knock-in mouse models of Huntington''s disease  [Striatum_mRNA] (SRP053398). The
data of this study is to analyze the expression profiles of RNA in mouse models 
with Hungtington's diseaase at different ages with different lengths of CAG 
repeats, to understand better the molecular changes related with this neurodegenarative 
disease.


```{r}
rse_gene_SRP053398 <- recount3::create_rse_manual(
  project = "SRP053398",
  project_home = "data_sources/sra",
  organism = "mouse",
  annotation = "gencode_v23",
  type = "gene")
```
#### Processing data

```{r}
# Transform the counts
assay(rse_gene_SRP053398, "counts") <- compute_read_counts(rse_gene_SRP053398)

# Before expanding the data we should explore the attributes
rse_gene_SRP053398$sra.sample_attributes[1:3]
```

```{r}
# Now we expand the attributes to process the data to make easier the analysis
rse_gene_SRP053398 <- expand_sra_attributes(rse_gene_SRP053398)
colData(rse_gene_SRP053398)[
    ,
    grepl("^sra_attribute", colnames(colData(rse_gene_SRP053398)))]
```


```{r}
# To perform the statistical analysis we need to use a correct R format
rse_gene_SRP053398$sra_attribute.age <- as.numeric(gsub(" month", "", rse_gene_SRP053398$sra_attribute.age))
rse_gene_SRP053398$sra_attribute.genotype <- factor(tolower(rse_gene_SRP053398$sra_attribute.genotype))
rse_gene_SRP053398$sra_attribute.Sex <- factor(rse_gene_SRP053398$sra_attribute.Sex)

# We can check the summary of the interest variables
summary(as.data.frame(colData(rse_gene_SRP053398)[
    ,
    grepl("^sra_attribute.[age|disease|RIN|sex]", colnames(colData(rse_gene_SRP053398)))
]))
```
We will check the quality of the data calculating the proportion of assigned genes.
```{r}
rse_gene_SRP053398$assigned_gene_prop <- rse_gene_SRP053398$recount_qc.gene_fc_count_all.assigned / 
  rse_gene_SRP053398$recount_qc.gene_fc_count_all.total
summary(rse_gene_SRP053398$assigned_gene_prop)
```
```{r}
# We can create another variable to make more general the analysis
rse_gene_SRP053398$state <- factor(ifelse(rse_gene_SRP053398$sra_attribute.genotype != 'wild type' , "knock-in", "wildtype"))
table(rse_gene_SRP053398$state)
```


```{r}
hist(rse_gene_SRP053398$assigned_gene_prop, main = "Assigned genes", xlab = "Proportion of assigned genes")
```
```{r}
# We can also check if there is difference in the quality of the data between the different states
with(colData(rse_gene_SRP053398), tapply(assigned_gene_prop, state, summary))
```
#### Normalization of the data

Normalizing data is essential because it ensures that all features contribute 
equally to the analysis, preventing bias towards variables with larger scales.
```{r}
# With the library of edgeR we will normalize the data
dge <- DGEList(
    counts = assay(rse_gene_SRP053398, "counts"),
    genes = rowData(rse_gene_SRP053398)
)
dge <- calcNormFactors(dge)
```
## Expression data analyisis

Here we will compare the differences between the different states of the mice (knock-in and wildtype),
and the differences of the knock-in mice CAG repeats lengths.

#### Data distribution plots

```{r}
ggplot(as.data.frame(colData(rse_gene_SRP053398)), aes(y = assigned_gene_prop, x = state,)) +
    geom_violin(trim = FALSE, alpha = 0.3, fill = 'red') +
    geom_boxplot(width = 0.15) +
    theme_bw(base_size = 20) +
    ylab("Assigned Gene Prop") +
    xlab("Age Group")
```
```{r}

ggplot(as.data.frame(colData(rse_gene_SRP053398)), aes(y = assigned_gene_prop, x = sra_attribute.genotype,
    fill = sra_attribute.genotype)) +
    geom_violin(trim = FALSE, alpha = 0.3, fill = 'red') +
    geom_boxplot(width = 0.15) +
    theme_bw(base_size = 20) +
    ylab("Assigned Gene Prop") +
    xlab("Age Group") + 
    theme(axis.text.x = element_blank())
  
```

#### Statistical model
```{r}
mod <- model.matrix(~ state + sra_attribute.Sex + sra_attribute.age + assigned_gene_prop,
    data = colData(rse_gene_SRP053398)
)
colnames(mod)
```

```{r}
mod2 <- model.matrix(~ sra_attribute.genotype + sra_attribute.Sex + sra_attribute.age + assigned_gene_prop,
    data = colData(rse_gene_SRP053398)
)
colnames(mod2)
```

Now with the statistical model, we can visualize the expression data between the 
two models.
```{r}
vGene <- voom(dge, mod, plot = TRUE)
```
```{r}
vGene2 <- voom(dge, mod2, plot = TRUE)
```

Lets see the data in an easier way.
```{r}
eb_results <- eBayes(lmFit(vGene))
eb_results2 <- eBayes(lmFit(vGene2))
```

```{r}
de_results <- topTable(
    eb_results,
    coef = 2,
    number = nrow(rse_gene_SRP053398),
    sort.by = "none"
)
dim(de_results)
```
```{r}
de_results2 <- topTable(
    eb_results2,
    coef = 2,
    number = nrow(rse_gene_SRP053398),
    sort.by = "none"
)
dim(de_results2)
```

```{r}
head(de_results)
```
```{r}
head(de_results2)
```


```{r}
# Differentially expressed genes
table(de_results$adj.P.Val < 0.05)
```
```{r}
# Differentially expressed genes
table(de_results2$adj.P.Val < 0.05)
```

```{r}
# Performing volcano plot
volcanoplot(eb_results, coef = 2, highlight = 3, names = de_results$gene_name)
```
```{r}
volcanoplot(eb_results2, coef = 2, highlight = 3, names = de_results$gene_name)
```


```{r}
# Performing cluster heatmap

# Extracting over/sub expressed genes
exprs_heatmap <- vGene$E[rank(de_results$adj.P.Val) <= 50, ]

df <- as.data.frame(colData(rse_gene_SRP053398)[, c("state", "sra_attribute.genotype", "sra_attribute.age", "sra_attribute.Sex")])
colnames(df) <- c("State", "gentoype", "Age", "Sex")

library("pheatmap")
pheatmap(
    exprs_heatmap,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    show_rownames = FALSE,
    show_colnames = FALSE,
    annotation_col = df
)
```
#### Multidimensional scaling
```{r}
col.group <- df$State
levels(col.group) <- brewer.pal(nlevels(col.group), "Set1")

col.group <- as.character(col.group)

plotMDS(vGene$E, labels = df$State, col = col.group)
```

```{r}
col.group2 <- df$gentoype
levels(col.group2) <- brewer.pal(nlevels(col.group2), "Set1")

col.group2 <- as.character(col.group2)

plotMDS(vGene2$E, labels = df$genotype, col = col.group2)
```


```{r}
col.group <- df$Age
levels(col.group) <- brewer.pal(nlevels(col.group), "Set1")

col.group <- as.character(col.group)

plotMDS(vGene$E, labels = df$Age, col = col.group)
```

## Biological discussion

The analysis of the data shows that there are differences in the expression of both models.
In the first model only comparing the state of the mice, there were more differential expressed genes
than the model that considers also the knock-in mice CAG repeats lengths. 
In the first model, there was a clear separation in the MDS Plot, that indicates that shows the effect
of the knock in. In the second model the effect of the CAG repeats lengths has an impact more specific 
in some genes like the Onecut1 or Klf7, that suggests that some genes are regulated in a dependant way
of the CAG repeats lengths.
In conclusion the knock-in the mutation of the Htt has a strong effect in gene expression, that could
be reflected in the modifications of multiple biological pathways, like inflammation, oxidative stress or
neurodegeneration. The CAG repeats lengths has a less impact in the gene expression, leading to a more
gradual or restricted effect in specific processes. These few overexpressed genes and subexpressed genes,
could be related to the severity of the disease. In the study already performed by Langelder et al. (2016), 
[@langfelder_2016] they mentioned that the CAG length and age are highly correlated with 13 striatal and 5 
cortical modules, implicated in the dysregulation of cyclic AMP signaling, cell death and procadherin genes.










